{\bf Practical Concurrency Testing} \\ \normalsize \vspace{1em} \begin{tabular}{c} {\em or, How I Learned to Stop Worrying and Love the Exponential Explosion} \end{tabular}

### 2018-08-01 thesis/tm.tex
few more tm runs

The notation {\tt testname}$(K,N)$ will denote a test configuration of $K$ threads, each running $N$ iterations of the test logic.
All tests were run on an 8-core 2.7GHz Core i7 with 32 GB RAM.
Reported CPU-times include time spent on all state spaces
Quicksand saw fit to run,
not just the maximal or the buggy state space;
for verification tests (run with {\tt -M}),
this still includes abandoned smaller jobs that were run to saturate the set of data-race preemption points
(Chapter~\ref{chap:quicksand}).
To minimize variance in CPU-time measurements,
I ensured the test machine was not loaded beyond normal web browser use,
and ran only one instance of Landslide at a time;
for further discussion of variance see \cite{sigbovik-htm}.

@@

			& 2,3 &\cpu{qqq}	& \ints{qqq}	& \cpu{qqq}	& \ints{qqq}	& \cpu{910.80}	& \ints{22409} \\

@@

			& 2,1 & \cpu{qqq}	& \ints{qqq}	& \cpu{qqq}	& \ints{qqq}	& \cpu{qqq}	& \ints{qqq} \\

@@

			& 2,4 & \cpu{qqq}	& \ints{qqq}	& \cpu{qqq}	& \ints{qqq}	& \cpu{qqq}	& \ints{qqq} \\

@@

			& 3,2 & \cpu{qqq}	& \ints{qqq}	& \cpu{qqq}	& \ints{qqq}	& \cpu{qqq}	& \ints{qqq} \\

@@

			& 2,2 & \cpu{qqq}	& \ints{qqq}	& \cpu{qqq}	& \ints{qqq}	& \cpu{qqq}	& \ints{qqq} \\

@@

			& 4,2 & \ETAdag{qqq} 	& \ETAdag{qqq} 	& \ETAdag{qqq}	& \ETAdag{todo atm}	& \cpu{3.28}	& \ints{1} \\
			%& 2,2 & \cpu{9.40}	& \ints{108}	& \cpu{8.92}	& \ints{76}	& \cpu{3.37}	& \ints{1} \\
			%& 3,1 & \cpu{10.30}	& \ints{132}	& \cpu{7.50}	& \ints{21}	& \cpu{3.64}	& \ints{1} \\
			& 2,1 & \cpu{545.45}	& \ints{15052}	& \cpu{244.20}	& \ints{6288}	& \cpu{109.84}	& \ints{2774} \\
			% (2,2) - bug found on regular w/ ?? ints, on retry sets w/ ?? ints, and codes w/ ?? ints
			{\tt avl\_fixed}
			& 2,1 & \cpu{642.73} 	& \ints{20416} 	& \cpu{293.70}	& \ints{9677}	& \cpu{108.69}	& \ints{2774} \\
			& 2,2 & \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq} \\
			& 3,1 & \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq} \\
			{\tt map\_basic}
			& 2,1 & \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq} \\
			{\tt map\_basicer}
			& 2,1 & \cpu{qqq} 	& \ints{qqq} 	& \cpu{qqq}	& \ints{qqq}	& \cpu{qqq}	& \ints{qqq} \\
			& 2,2 & \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq} \\
			& 3,1 & \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq} \\

@@

It is difficult to discern from prior work a concrete standard
for what values of $(K, N)$ constitute a ``good'' degree of verification.
One recent paper \cite{tsopso},
which likewise extended DPOR with a new dimension of concurrency (weak memory orderings)
reported verifying programs with up to 10 concurrent events,
presumably shared memory accesses.
Another \cite{optimal-dpor}
reported test cases with as many as 19 threads,
although with what must be very little synchronization or memory conflicts,
%although with what nature of synchronization or memory conflicts is not clear,
as even their baseline DPOR checked only 4096 interleavings on that test.
%
Table~\ref{tab:tm-test-case-evence} confers a sense of the complexity of this evaluation's test suite,
with the final column showing the approximate maximum number of preemption points
reached among Landslide's verifications
(i.e., $({\sf txn}+{\sf sync}+{\sf race})*K*N$ for the highest $(K,N)$ completed.
%%%% on second thought, i'm not gonna do this -- any branch that xend shows up in, is gonna lose any of the pps coming
%%%% from the failure path's mxes or data races, so xend will never be an *additional* pp
%%%% and of course it will scare the committee
% \footnote{The 1.5 multiplier on {\sf txn} represents the fact that for each transaction,
% the {\tt xend} preemption point marking the end of its success path shows up about half the time,
% the other half being failure injections.
% % in reality it's usually less than half, because (1) drpps can show up in failure paths, which will get interleaved around,
% % thereby giving the abort path more interleavings, and (2) under -A -S, you might get multiple codes for duplicate abort paths
% % abort sets could theoretically push them above half but probably not in any test case here
% }).

\begin{table}[h]
	\begin{center}
		\begin{tabular}{c||c|c|c||c}
			{\bf test} & {\bf \#txn} & {\bf \#sync} & {\bf \#race} & {\bf max events verified} \\
			\hline
			{\tt htm2}		& 1	& 2	& 4	& 42	\\ % 2,3
			{\tt counter}		& 1	& 0	& 1	& 16	\\ % 2,4
			{\tt swap}		& 1	& 4	& 8	& 52	\\ % 2,2
			{\tt fig63}		& 1	& 0	& $K$	& 24	\\ % 3,2 k=3 -> 24; 4,2 -> 40 (not yet); 2,4 -> 24
			{\tt avl\_insert}	& 1	& 2	& 7	& 20	\\ % 2,1 nb. only 5 drpps under -A -S but w/e
			{\tt avl\_fixed}	& 1	& 2	& 9	& 24	\\ % 2,1 nb. again, only 5 under -A -S
			% the extra 2 over basicer is from resize, which only 1 thread does, but K*N is 0 here anyway so vov
			{\tt map\_basic}	& 1	& 4	& ?	& 0	\\ % lol
			{\tt map\_basicer}	& 1	& 2	& ?	& ?	\\
		\end{tabular}
	\end{center}
	\caption{Number of concurrency events per iteration of each test case.
	Note that no test used any synchronization besides mutexes % ;
	%while they do use the P2 thread API, %in the first place,
	(the P2 thread API was annotated as trusted (\sect{\ref{sec:landslide-config-landslide}})
	and so does not contribute to state space size).
	Also note that ``\#race'' means the number of unique accesses identified as racy,
	rather than racing pairs
	(the other half of a pair might well be within a transaction, which cannot be preempted on).
	}
	\label{tab:tm-test-case-evence}
\end{table}

In the case of {\tt htm2},
it is easy to look at the program with human intuition and judge that,
because the protocol's only state is stored in a single boolean,
with no unbounded-capacity data structures or
contention-dependent exponential backoff loops,
it would be unimaginable that adding a 5th thread to the system
could make any difference in correctness where 4 threads could not,
or that a 4th repetition between 2 threads could make a difference where 3 could not,
and ultimately that it must safely generalize to all $(K, N)$.
% under certain invisible assumptions like the system runs with SC memory! (spoilers!!)
Generalizing the verification is not so straightforward
for more complicated algorithms,
which may involve complex conflict patterns such as tree rebalancing or map resizing;
other formal verification approaches aside,
the user must ultimately be content with the probabilistic assurance that
as verified $K$ and $N$ increase,
the likelihood that a bug exists which requires more threads or iterations to expose grows ever lower
% (in the author's fancy, it may as well drop exponentially in the same way as the state spaces grow)
(for example, none of the bugs in \sect{\ref{sec:tm-eval-bugs}} required any higher parameters than $(2,2)$ to expose).
% and by analogy with ICB, none of the bugs in quicksand evaluation took more than a preemption bound of 3 to expose
% but it's not clear how to state this "analogy" concisely in text

@@

>>>>

The notation {\tt testname}$(K,N)$ will denote a test configuration of $K$ threads, each running $N$ iterations of the test logic.
All tests were run on an 8-core 2.7GHz Core i7 with 32 GB RAM.
Reported CPU-times include time spent on all state spaces
Quicksand saw fit to run,
not just the maximal or the buggy state space;
for verification tests (run with {\tt -M}),
this still includes abandoned smaller jobs that were run to saturate the set of data-race preemption points
(Chapter~\ref{chap:quicksand}).
To minimize variance in CPU-time measurements,
I ensured the test machine was not loaded beyond normal web browser use,
and ran only one instance of Landslide at a time;
for further discussion of variance see \cite{sigbovik-htm}.
The number of interleavings in each state space are, of course, deterministic and do not vary across runs.

@@

			& 2,3 &\cpu{3923.32}	& \ints{104914}	& \cpu{230.71}	& \ints{7528}	& \cpu{910.80}	& \ints{22409} \\

@@

			& 2,1 & \cpu{6.90}	& \ints{10}	& \cpu{7.04}	& \ints{8}	& \cpu{3.35}	& \ints{4} \\

@@

			& 2,4 & \cpu{3433.08}	& \ints{86950}	& \cpu{1134.62}	& \ints{25398}	& \cpu{752.51}	& \ints{19128} \\

@@

			& 3,2 & \cpu{2492.17}	& \ints{60606}	& \cpu{2040.98}	& \ints{44862}	& \cpu{565.97}	& \ints{14304} \\

@@

			& 2,2 & \cpu{qqq}	& \ints{qqq}	& \cpu{497.11}	& \ints{11427}	& \cpu{461.54}	& \ints{9046} \\

@@

			& 4,2 & \ETAdag{qqq} 	& \ETAdag{qqq} 	& \ETAdag{22d 17h}	& \ETAdag{3129034}	& \cpu{3.28}	& \ints{1} \\
			%[JOB 1] progress: 827215/3129034 brs (26.436748%), ETA 22d 7h 18m 34s (elapsed 9h 59m 56s)

			% measured with !CONSIDER_ONLY_MOST_RECENT_DPOR_PREFERRED_TID
			%& 2,2 & \cpu{9.40}	& \ints{108}	& \cpu{8.92}	& \ints{76}	& \cpu{3.37}	& \ints{1} \\
			%& 3,1 & \cpu{10.30}	& \ints{132}	& \cpu{7.50}	& \ints{21}	& \cpu{3.64}	& \ints{1} \\
			\hline
			{\tt avl\_insert}
			& 2,1 & \cpu{545.45}	& \ints{15052}	& \cpu{244.20}	& \ints{6288}	& \cpu{109.84}	& \ints{2774} \\
			% (2,2) - bug found on regular w/ ?? ints, on retry sets w/ ?? ints, and codes w/ ?? ints
			\hline
			{\tt avl\_fixed}
			& 2,1 & \cpu{642.73} 	& \ints{20416} 	& \cpu{293.70}	& \ints{9677}	& \cpu{108.69}	& \ints{2774} \\
			& 2,2 & \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq} \\
			& 3,1 & \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq} \\
			\hline
			{\tt map\_basic}
			& 2,1 & \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq} \\
			\hline
			{\tt map\_basicer}
			& 2,1 & \cpu{qqq} 	& \ints{qqq} 	& \cpu{qqq}	& \ints{qqq}	& \cpu{qqq}	& \ints{qqq} \\
			& 2,2 & \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq} \\
			& 3,1 & \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq} \\

@@

It is difficult to discern from prior work a concrete standard
for what values of $(K, N)$ constitute a ``good'' degree of verification.
One recent paper \cite{tsopso},
which likewise extended DPOR with a new dimension of concurrency (weak memory orderings)
reported verifying programs with up to 10 concurrent events,
presumably shared memory accesses.
Another \cite{optimal-dpor}
reported test cases with as many as 19 threads,
although with what must be very little synchronization or memory conflicts,
%although with what nature of synchronization or memory conflicts is not clear,
as even their baseline DPOR checked only 4096 interleavings on that test.
%
Table~\ref{tab:tm-test-case-evence} confers a sense of the complexity of this evaluation's test suite,
with the final column showing the approximate maximum number of preemption points
reached among Landslide's verifications
(i.e., $({\sf txn}+{\sf sync}+{\sf race}) \times K \times N$ for the highest $K \times N$ completed.
%%%% on second thought, i'm not gonna do this -- any branch that xend shows up in, is gonna lose any of the pps coming
%%%% from the failure path's mxes or data races, so xend will never be an *additional* pp
%%%% and of course it will scare the committee
% \footnote{The 1.5 multiplier on {\sf txn} represents the fact that for each transaction,
% the {\tt xend} preemption point marking the end of its success path shows up about half the time,
% the other half being failure injections.
% % in reality it's usually less than half, because (1) drpps can show up in failure paths, which will get interleaved around,
% % thereby giving the abort path more interleavings, and (2) under -A -S, you might get multiple codes for duplicate abort paths
% % abort sets could theoretically push them above half but probably not in any test case here
% }).

\begin{table}[h]
	\begin{center}
		\small
		\begin{tabular}{c||c|c|c||c}
			{\bf test} & {\bf \#txn} & {\bf \#sync} & {\bf \#race} & {\bf max events verified} \\
			\hline
			{\tt htm2}		& 1	& 2	& 4	& 42	\\ % 2,3
			{\tt counter}		& 1	& 0	& 1	& 16	\\ % 2,4
			{\tt swap}		& 1	& 4	& 8	& 52	\\ % 2,2
			{\tt fig63}		& 1	& 0	& $K$	& 24	\\ % 3,2 k=3 -> 24; 4,2 -> 40 (not yet); 2,4 -> 24
			{\tt avl\_insert}	& 1	& 2	& 7	& 20	\\ % 2,1 nb. only 5 drpps under -A -S but w/e
			{\tt avl\_fixed}	& 1	& 2	& 9	& 24	\\ % 2,1 nb. again, only 5 under -A -S
			% the extra 2 over basicer is from resize, which only 1 thread does, but K*N is 0 here anyway so vov
			{\tt map\_basic}	& 1	& 4	& ?	& 0	\\ % lol
			{\tt map\_basicer}	& 1	& 2	& ?	& ?	\\
		\end{tabular}
	\end{center}
	\caption{Number of concurrency events per iteration of each test case.
	Note that no test used any synchronization besides mutexes % ;
	%while they do use the P2 thread API, %in the first place,
	(the P2 thread API was annotated as trusted (\sect{\ref{sec:landslide-config-landslide}})
	and so does not contribute to state space size).
	Also note that ``\#race'' means the number of unique accesses identified as racy,
	rather than racing pairs
	(the other half of a pair might well be within a transaction, which cannot be preempted on).
	}
	\label{tab:tm-test-case-evence}
\end{table}

In the case of {\tt htm2},
it is easy to look at the program with human intuition and judge that,
because the protocol's only state is stored in a single boolean,
with no unbounded-capacity data structures or
contention-dependent exponential backoff loops,
it would be unimaginable that adding a 5th thread to the system
could make any difference in correctness where 4 threads could not,
or that a 4th repetition between 2 threads could make a difference where 3 could not,
and ultimately that it must safely generalize to all $(K, N)$.
% under certain invisible assumptions like the system runs with SC memory! (spoilers!!)
Generalizing the verification is not so straightforward
for more complicated algorithms,
which may involve complex conflict patterns such as tree rebalancing or map resizing;
other formal verification approaches aside,
the user must ultimately be content with the probabilistic assurance that
as verified $K$ and $N$ increase,
the likelihood that a bug exists which requires more threads or iterations to expose grows ever lower
% (in the author's fancy, it may as well drop exponentially in the same way as the state spaces grow)
(for example, none of the bugs in \sect{\ref{sec:tm-eval-bugs}} required any higher parameters than (2,2) to expose).
% and by analogy with ICB, none of the bugs in quicksand evaluation took more than a preemption bound of 3 to expose
% but it's not clear how to state this "analogy" concisely in text

@@

As an example, note how STM semantics reduced {\tt fig63}'s state space on all $(K,N)$ configurations to 1.
Because its transactional paths are all mutually independent,
DPOR identifies no need either to inject conflict aborts or to reorder threads.
However, this skips the slow-path consistency assertion completely,
which the programmer may have intended to run ``every so often''
